---
title: "Lipidome FA analysis"
author: "Bryan Gonzales"
date: "2024-04-11"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is the analysis of the lipidomics data after factor analysis

```{r cars}
prot <- read.csv("CFA_lipidome_all.csv", header = TRUE, sep = ";")
head(prot)

library(dplyr)
prot <- prot %>% filter(x == "")
table(prot$Group, prot$timepoint)

```

Perform kwash vs others analysis using conditional logistic regression at admission to determine which LV are interesting

```{r}
library(survival)
library(dplyr)

admission <- prot %>% filter(timepoint == "A")
admission$Kall <- ifelse(admission$Group == "Kwash", 1, 0)
table(admission$Kall, admission$Group)

fit <- clogit(Kall ~ lLV5 + strata(Pair), data = admission)
summary(fit)

```
Loop to do clogit in all LV's, extract estimate and FDR p-value

```{r}

GLM.run<-function(y) {
  form <- as.formula(paste0("Kall ~ strata(Pair) +", y))
  fit<-(clogit(form, data=admission))}

GLMs.out <- lapply(colnames(admission[,c(12:32)]),GLM.run )
results<-lapply(GLMs.out, function(x){summary(x)})
results

#extracting pValues and FDR correction
results<-lapply(GLMs.out, function(x){coef(summary(x))[1,5]})
pVal = as.matrix(results)
rownames(pVal) = colnames(admission[12:32])
fdr = as.matrix(p.adjust(as.vector(results), method ="fdr", n=21)) #n=number of metabolites
pVal=cbind(pVal, fdr)
colnames(pVal) = c("pVal", "fdr p")
pVal <- as.matrix(pVal)
View(as.data.frame(pVal))

#extracting estimate
results_est<-lapply(GLMs.out, function(x){coef(summary(x))[1,"exp(coef)"]}) #[x,4] change x to number of factors + 1
estimate = as.matrix(results_est)
rownames(estimate) = colnames(admission[12:32])
View(as.data.frame(estimate))

#extracting std error
results_se<-lapply(GLMs.out, function(x){coef(summary(x))[1,"se(coef)"]}) #[x,4] change x to number of factors + 1
stderror = as.matrix(results_se)
rownames(stderror) = colnames(admission[12:32])
View(as.data.frame(stderror))

#compile statistics table
uni_stat <- cbind(estimate, stderror, pVal)
colnames(uni_stat) <- c("exp(estimate)", "std error", "pVal", "FDRpVal")
View(as.data.frame(uni_stat))
write.csv(uni_stat, file = "admission_clogit_matchstrata.csv")
uni_stat

```
```{r}

variable_names <- paste0("lLV", 1:21)

# Create a boxplot for each variable in the list
library(gridExtra)
library(ggplot2)

variable_names <- colnames(admission)[-c(1:11,33)] # Exclude the 'Group' column

# Create individual boxplots
plot_list <- lapply(variable_names, function(variable) {
  ggplot(admission, aes_string(x = "Group", y = variable)) +
    geom_boxplot(aes(fill = Group), color = "black", outlier.shape = 21, outlier.size = 2, width = 0.5) +
    labs(title = variable, x = "Group", y = variable) +
    theme_minimal(base_size = 15) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          legend.position = "none") +
    scale_fill_brewer(palette = "Set3")
})

# Arrange the boxplots in a 3x3 grid
grid.arrange(grobs = plot_list, ncol = 8)



```
significant lLVs are: lLV 1, 2, 5, 10, 13:16 and 20 (9 lLVs)

Second assessment, lLV that is different in kwash should be associated with severity of oedema

```{r}

kadmission <- admission %>% filter(Group == "Kwash")
head(kadmission)

kadmission$edema_s[kadmission$edema_s == "+"] <- 1
kadmission$edema_s[kadmission$edema_s == "++"] <- 2
kadmission$edema_s[kadmission$edema_s == "+++"] <- 3
kadmission$edema_s <- as.factor(kadmission$edema_s)

kadmission <- kadmission[,-c(14:15, 17:20, 22:23, 28:30, 32:33)]

library(MASS)

GLM.run<-function(y) {
  form <- as.formula(paste0("edema_s ~ ", y))
  fit<-( polr(form, data=kadmission, Hess = TRUE))}

GLMs.out <- lapply(colnames(kadmission[,c(12:20)]),GLM.run )
results<-lapply(GLMs.out, function(x){summary(x)})
results

#extracting tValues
results<-lapply(GLMs.out, function(x){coef(summary(x))[1,"t value"]})
tVal = as.matrix(results)
rownames(tVal) = colnames(kadmission[12:20])

#extracting estimate
results_est<-lapply(GLMs.out, function(x){coef(summary(x))[1,"Value"]}) #[x,4] change x to number of factors + 1
estimate = as.matrix(results_est)
rownames(estimate) = colnames(admission[12:20])
View(as.data.frame(estimate))

#compile statistics table
uni_stat <- as.matrix(cbind(estimate, tVal))
colnames(uni_stat) <- c("estimate", "tVal")
write.csv(uni_stat, file = "edemaGrade_polr_matchstrata.csv", row.names = F)

#calculate pvalue

polr <- read.csv("edemaGrade_polr_matchstrata.csv", header = T, sep = ",")
polr$p_value <- pnorm(abs(polr$tVal), lower.tail = FALSE) * 2
polr$fdr = p.adjust(polr$p_value, method ="fdr", n=9)
polr
rownames(polr) = colnames(kadmission[12:20])
write.csv(polr, file = "edemaGrade_polr_matchstrata.csv")

```
```{r}

# Create a boxplot for each variable in the list
library(gridExtra)
library(ggplot2)

variable_names <- colnames(kadmission)[-c(1:11)] # Exclude the 'Group' column

# Create individual boxplots
plot_list <- lapply(variable_names, function(variable) {
  ggplot(admission, aes_string(x = "edema_s", y = variable)) +
    geom_boxplot(aes(fill = edema_s), color = "black", outlier.shape = 21, outlier.size = 2, width = 0.5) +
    labs(x = "", y = variable) +
    theme_minimal(base_size = 15) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          legend.position = "none") +
    scale_fill_brewer(palette = "Set3")
})

# Arrange the boxplots in a 3x3 grid
grid.arrange(grobs = plot_list, ncol = 3)


```
Next, it should resolve over time

```{r}

res <- prot%>% filter(Group == "Kwash")

boxplot(lLV7 ~ timepoint, data = res)


library(lmerTest)

resolvelLV5 <- lmer(lLV5 ~ timepoint + age + sex + (1 | subjid), data = res)
summary(resolvelLV5)


resolvelLV16 <- lmer(lLV16 ~ timepoint + age + sex + (1 | subjid), data = res)
summary(resolvelLV16)


```
```{r}

boxplot(lLV16 ~ timepoint*Group, data = prot)

ggplot(prot, aes(x = interaction(timepoint, Group), y = lLV16)) +
    geom_boxplot(aes(fill = Group), color = "black", outlier.shape = 21, outlier.size = 2, width = 0.5,
                 position = position_dodge(width = 1)) +
    labs(title = "lLV5", x = "Follow-up", y = "lLV5") +
    theme_minimal(base_size = 15) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          legend.position = "none") +
    scale_fill_brewer(palette = "Set3")

```


Sensitivity analysis to test lLV5

Remove malaria patients

```{r}

malaria <- c("P010", "P014", "P017", "P022", "P025", "P028", 
                    "P031", "P034", "P035", "P038", "P046", "P047", 
                    "P056", "P059", "P070", "P082", "P095", "P140", 
                    "P167", "P170", "P179", "P182", "P184", "P187", 
                    "P188", "PR02", "PR04")

sensitivity <- admission %>% 
  filter(!subjid %in% malaria)

table(sensitivity$Group, sensitivity$timepoint)


#different in kwashiorkor?

fit <- clogit(Kall ~ lLV5 + strata(Pair), data = sensitivity)
summary(fit)

#plot

 ggplot(sensitivity, aes_string(x = "Group", y = "lLV5")) +
    geom_boxplot(aes(fill = Group), color = "black", outlier.shape = 21, outlier.size = 2, width = 0.5) +
    labs(title = "lLV5", x = "Group", y = "lLV5") +
    theme_minimal(base_size = 15) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          legend.position = "none") +
    scale_fill_brewer(palette = "Set3")

#associated with oedema severity?

sensitivity <- kadmission %>% 
  filter(!sample %in% malaria)

table(sensitivity$edema_s)

ol = polr(edema_s ~ lLV5, data=sensitivity, Hess = TRUE)
summary(ol)
(ctable <- coef(summary(ol)))
pval <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
(ctable <- cbind(ctable, "p value" = pval))

#plot

ggplot(sensitivity, aes_string(x = "edema_s", y = "lLV5")) +
    geom_boxplot(aes(fill = edema_s), color = "black", outlier.shape = 21, outlier.size = 2, width = 0.5) +
    labs(title = "lLV5", x = "Edema severity", y = "lLV5") +
    theme_minimal(base_size = 15) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          legend.position = "none") +
    scale_fill_brewer(palette = "Set3")

#resolve over time in kwashiorkor?

LV5 <- prot[,c(1:11,16)]
LV5k <- LV5%>% filter(Group == "Kwash")


sensitivity <- LV5k %>% 
  filter(!sample %in% malaria)

table(sensitivity$timepoint)


resolveS <- lmer(lLV5 ~ timepoint + age + sex + (1 | subjid), data = sensitivity)
summary(resolveS)


ggplot(sensitivity, aes(x = interaction(timepoint, Group), y = lLV5)) +
    geom_boxplot(aes(fill = Group), color = "black", outlier.shape = 21, outlier.size = 2, width = 0.5,
                 position = position_dodge(width = 1)) +
    labs(title = "lLV5", x = "Follow-up", y = "lLV5") +
    theme_minimal(base_size = 15) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          legend.position = "none") +
    scale_fill_brewer(palette = "Set3")


```
Only include kwashiorkor with negative urine protein and urine blood

```{r}
uprotsang <- c("P001A", "P004A", "P010A", "P034A", "P037A", "P043A", "P046A", "P052A", "P058A", "P061A", "P067A", 
               "P073A", "P091A", "P097A", "P106A", "P109A", "P127A", "P145A", "P157A", "P160A", "P163A", "P166A", "P169A", "P175A", 
               "P178A", "P184A", "P187A", "P190A", "P193A", "P196A", "P199A", "P202A")

sensitivity <- admission %>% 
  filter(!sample %in% uprotsang)

table(sensitivity$Group)

#different in kwashiorkor?

fit <- clogit(Kall ~ lLV5 + strata(Pair), data = sensitivity)
summary(fit)


ggplot(sensitivity, aes_string(x = "Group", y = "lLV5")) +
    geom_boxplot(aes(fill = Group), color = "black", outlier.shape = 21, outlier.size = 2, width = 0.5) +
    labs(title = "lLV5", x = "Group", y = "lLV5") +
    theme_minimal(base_size = 15) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          legend.position = "none") +
    scale_fill_brewer(palette = "Set3")



#associated with oedema severity?

sensitivity <- kadmission %>% 
  filter(!sample %in% uprotsang)

ol = polr(as.factor(edema_s) ~ lLV5 + age + sex, data=sensitivity, Hess = TRUE)
summary(ol)
(ctable <- coef(summary(ol)))
pval <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
(ctable <- cbind(ctable, "p value" = pval))

table(sensitivity$edema_s)

ggplot(sensitivity, aes_string(x = "edema_s", y = "lLV5")) +
    geom_boxplot(aes(fill = edema_s), color = "black", outlier.shape = 21, outlier.size = 2, width = 0.5) +
    labs(title = "lLV5", x = "Edema severity", y = "lLV5") +
    theme_minimal(base_size = 15) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          legend.position = "none") +
    scale_fill_brewer(palette = "Set3")


#resolve over time in kwashiorkor?

LV7 <- prot[,c(1:11,16)]
LV7k <- LV7%>% filter(Group == "Kwash")


sensitivity <- LV7k %>% 
  filter(!sample %in% uprotsang)


resolveS <- lmer(lLV5 ~ timepoint + age + sex + (1 | subjid), data = sensitivity)
summary(resolveS)


ggplot(sensitivity, aes(x = interaction(timepoint, Group), y = lLV5)) +
    geom_boxplot(aes(fill = Group), color = "black", outlier.shape = 21, outlier.size = 2, width = 0.5,
                 position = position_dodge(width = 1)) +
    labs(title = "lLV5", x = "Follow-up", y = "lLV5") +
    theme_minimal(base_size = 15) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          legend.position = "none") +
    scale_fill_brewer(palette = "Set3")

```

