---
title: "Proteome FA analysis"
author: "Bryan Gonzales"
date: "2024-04-11"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is the unvariable analysis of the proteomics data

```{r cars}
prot <- read.csv("proteomics_processed.csv", header = TRUE, sep = ",")
head(prot)
```

Perform kwash vs others analysis using conditional logistic regression at admission to determine which LV are interesting

```{r}
library(survival)
library(dplyr)

admission <- prot %>% filter(timepoint == "A")
admission$Kall <- ifelse(admission$Group == "Kwash", 1, 0)
table(admission$Kall, admission$Group)

```
Loop to do clogit in all LV's, extract estimate and FDR p-value

```{r}

GLM.run<-function(y) {
  form <- as.formula(paste0("Kall ~ strata(Pair) +", y))
  fit<-(clogit(form, data=admission))}

GLMs.out <- lapply(colnames(admission[,c(12:319)]),GLM.run )
results<-lapply(GLMs.out, function(x){summary(x)})
results

#extracting pValues and FDR correction
results<-lapply(GLMs.out, function(x){coef(summary(x))[1,5]})
pVal = as.matrix(results)
rownames(pVal) = colnames(admission[12:319])
fdr = as.matrix(p.adjust(as.vector(results), method ="fdr", n=308)) #n=number of metabolites
pVal=cbind(pVal, fdr)
colnames(pVal) = c("pVal", "fdr p")
pVal <- as.matrix(pVal)
View(as.data.frame(pVal))

#extracting estimate
results_est<-lapply(GLMs.out, function(x){coef(summary(x))[1,"exp(coef)"]}) #[x,4] change x to number of factors + 1
estimate = as.matrix(results_est)
rownames(estimate) = colnames(admission[12:319])
View(as.data.frame(estimate))

#extracting std error
results_se<-lapply(GLMs.out, function(x){coef(summary(x))[1,"se(coef)"]}) #[x,4] change x to number of factors + 1
stderror = as.matrix(results_se)
rownames(stderror) = colnames(admission[12:319])
View(as.data.frame(stderror))

#compile statistics table
uni_stat <- cbind(estimate, stderror, pVal)
colnames(uni_stat) <- c("exp(estimate)", "std error", "pVal", "FDRpVal")
View(as.data.frame(uni_stat))
write.csv(uni_stat, file = "admission_clogit_matchstrata.csv")

```

```{r}

# Create a volcano plot for each variable in the list
library(ggplot2)

uni_stat <- as.data.frame(uni_stat)
uni_stat$nlp <- -log10(as.numeric(uni_stat$FDRpVal))
uni_stat$est <- log10(as.numeric(uni_stat$`exp(estimate)`))

ggplot(uni_stat, aes(x = est, y = nlp)) +
  geom_point(aes(color = ifelse(nlp > 1.3, "red", "grey")), size = 3) +
  scale_color_identity() +  # Use the color column directly
  theme_minimal() +
  labs(title = "Volcano Plot", x = "Estimate (Log OR)", y = "-log10(p-value)") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +  # Horizontal line for p-value threshold
  geom_vline(xintercept = c(0,0), linetype = "dashed", color = "blue")       # Vertical lines for estimate thresholds  

#create boxplot for weird protein

boxplot(Prot183 ~ Group, admission)

```

Select variables with FDRpVal < 0.05

Second assessment, protein should be associated with severity of oedema

```{r}

kadmission <- admission %>% filter(Group == "Kwash")
head(kadmission)

kadmission$edema_s[kadmission$edema_s == "+"] <- 1
kadmission$edema_s[kadmission$edema_s == "++"] <- 2
kadmission$edema_s[kadmission$edema_s == "+++"] <- 3
kadmission$edema_s <- as.factor(kadmission$edema_s)

#select relevant proteins

round2 <- rownames(uni_stat %>% filter(FDRpVal < 0.05))
round2 <- cbind(kadmission[, 1:11], kadmission[,round2])
View(round2)

library(MASS)

ol = polr(edema_s ~ Prot183, data=kadmission, Hess = TRUE)
summary(ol)
(ctable <- coef(summary(ol)))
pval <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
(ctable <- cbind(ctable, "p value" = pval))


GLM.run<-function(y) {
  form <- as.formula(paste0("edema_s ~ ", y))
  fit<-( polr(form, data=round2, Hess = TRUE))}

GLMs.out <- lapply(colnames(round2[,c(12:176)]),GLM.run )
results<-lapply(GLMs.out, function(x){summary(x)})
results

#extracting tValues
results<-lapply(GLMs.out, function(x){coef(summary(x))[1,"t value"]})
tVal = as.matrix(results)
rownames(tVal) = colnames(kadmission[12:176])

#extracting estimate
results_est<-lapply(GLMs.out, function(x){coef(summary(x))[1,"Value"]}) #[x,4] change x to number of factors + 1
estimate = as.matrix(results_est)
rownames(estimate) = colnames(admission[12:176])
View(as.data.frame(estimate))

#compile statistics table
uni_stat <- as.matrix(cbind(estimate, tVal))
colnames(uni_stat) <- c("estimate", "tVal")
write.csv(uni_stat, file = "edemaGrade_polr_matchstrata.csv", row.names = F)

#calculate pvalue

polr <- read.csv("edemaGrade_polr_matchstrata.csv", header = T, sep = ",")
polr$p_value <- pnorm(abs(polr$tVal), lower.tail = FALSE) * 2
polr$fdr = p.adjust(polr$p_value, method ="fdr", n=165)
polr
rownames(polr) = colnames(admission[12:176])
write.csv(polr, file = "edemaGrade_polr_matchstrata.csv")

```
No individual protein passes the test

```{r}

#Manhattan plot

n = rep(1:nrow(polr))

ggplot(polr, aes(x = n, y = -log10(fdr))) +
  geom_point(aes(color = ifelse("grey")), size = 3) +
  labs(title = "Manhattan Plot",
       x = "Protein",
       y = expression(-log[10](italic(P)))) +
  theme_minimal() +
  theme(legend.position = "none") + 
  geom_hline(yintercept = 1.3, linetype = "dashed", color = "red") + 
  coord_cartesian(ylim = c(0, 2))

```
```

Perform PCA

```{r}

# Perform PCA
pca_result <- prcomp(prot[,-(1:11)], center = TRUE, scale. = TRUE)

library(ggplot2)
library(ggbiplot)

# Prepare the data for plotting
pca_data <- as.data.frame(pca_result$x)
pca_data$group <- prot$Group

explained_variance <- pca_result$sdev^2 / sum(pca_result$sdev^2) * 100
x_label <- paste0("PC1 (", round(explained_variance[1], 2), "%)")
y_label <- paste0("PC2 (", round(explained_variance[2], 2), "%)")

# Plot the PCA results using ggplot2
ggplot(pca_data, aes(x = PC1, y = PC2, color = group, label = group)) +
  geom_point(aes(color = group), size = 5) +
  #geom_text(vjust = 1.5, hjust = 1.5) +
  theme_minimal() +
  labs(title = "PCA Plot", x = x_label, y = y_label) +
  theme(legend.position = "right")

```


