---
title: "Urine metabolomics analysis using FAD"
author: "Bryan Gonzales"
date: "2024-04-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is the analysis of urine metabolomics data using factor analysis approach. The positive and negative modes have been combined using MSCombine.

```{r}

data <- read.csv("MSCombine_urine_metabolomics.csv", header = T, sep = ",")
colnames(data)[1:15]

```
Perform FAD analysis on combined positive and negative MS modes

```{r}
#first remove the demographic data

meta <- as.matrix(data[,-c(1:9)])

library(fad)

BICs <- rep(0,101) # first optimize factors ranging from 1 to 100

for(i in 1:101){
  out = fad(x = meta,factors = i-1)
  BICs[i] = out$BIC
}

BICs
plot(BICs)

ind = which.min(BICs) # obtain the optimal factor favored with lowest BIC 

ind

```

Run FAD using the number of latent variables with the lowest BIC.

```{r}
out1 = fad(x = meta, factors = 8, scores = c("Bartlett"), rotation = "varimax") ## X is the data
eig <- data.frame(out1$scores)
head(eig)

demog <- data[,1:9]
kdat <- cbind(demog, eig)
head(kdat)

write.csv(kdat, "factor_scores.csv", row.names = F)

#try out some LVs
par(mfrow = c(2, 4))
boxplot(X1 ~ Group, data = kdat)
boxplot(X2 ~ Group, data = kdat)
boxplot(X3 ~ Group, data = kdat)
boxplot(X4 ~ Group, data = kdat)
boxplot(X5 ~ Group, data = kdat)
boxplot(X6 ~ Group, data = kdat)
boxplot(X7 ~ Group, data = kdat)
boxplot(X8 ~ Group, data = kdat)

par(mfrow = c(1, 1))
```

Extract loadings
```{r}
load_cut <- data.matrix(out1$loadings)
head(load_cut)

rownames(load_cut) <- colnames(meta)

```


We impose a factor loading cut-off. we will use 0.49. This means that proteins that have less than 0.4 loadings to the latent variable will be removed - leading behind only those proteins that are contributing highly to the latent variable.

We then perform confirmatory factor analysis using lavaan using only strongly associated proteins per latent variable.

Then, we extract a new scores matrix - the predicted factor so that each observation gets a value for each latent variable.

```{r}

load_cut[load_cut < abs(0.49)] <- 0
head(load_cut)

write.csv(load_cut, "loadings_metabolomeFAD.csv")
```

Extract scores using the load_cut

```{r}
library(psych)

fxad <- factor.scores(meta, load_cut, Phi = NULL, method = c("Bartlett"),rho=NULL,impute="none")

scores_ad <- fxad$scores
head(scores_ad)

CFA_proteome_admission <- cbind(demog, scores_ad)

colnames(CFA_proteome_admission)[10:17] <- c("uLV1", "uLV2", "uLV3", "uLV4", "uLV5", "uLV6", "uLV7", "uLV8")
head(CFA_proteome_admission)

write.csv(CFA_proteome_admission, file= "FAD_urine_metabolome.csv", row.names=F)

```

visualise network including clinical biochem using partial correlations

```{r}

clin <- read.csv("urine_clinicalBiochem.csv", header = T, sep = ",")
head(clin)

metclin <- merge(clin[,-c(2:5)], CFA_proteome_admission, by = "subjid")

write.csv(metclin, file= "FAD_urine_metabolome_clinical.csv", row.names=F)
head(metclin)
```

Perform partial correlation

```{r}

library(psych)
library(qgraph)
library(glasso)
# Load example data (replace with your own dataset)
colnames(metclin)
ADclass2 <- metclin[,-c(1,8:15)]

# Calculate partial correlations
partial_cor <- cor_auto(ADclass2)

# Create network plot
qgraph(partial_cor, layout = "spring", minimum = 0.3)

sparse_partial_cor <- glasso(partial_cor, rho = 0.1)

# Create network plot
qgraph(sparse_partial_cor, layout = "spring", minimum = 0.2, labels = colnames(ADclass2))
qgraph(sparse_partial_cor, layout = "circle", minimum = 0.2, labels = colnames(ADclass2))
```
Visualize clin

```{r}

par(mfrow = c(2, 3))
boxplot(log(K) ~ group, data = clin)
boxplot(log(Na) ~ group, data = clin)
boxplot(log(Cl) ~ group, data = clin)
boxplot(log(Alb) ~ group, data = clin)
boxplot(log(Urea) ~ group, data = clin)
boxplot(log(Crea) ~ group, data = clin)

par(mfrow = c(1, 1))

```


