---
title: "Plasma Metabolome FA analysis"
author: "Bryan Gonzales"
date: "2024-04-11"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is the analysis of the plasma metabolomics data after factor analysis

```{r cars}
prot <- read.csv("CFA_metabolome_all.csv", header = TRUE, sep = ",")
head(prot)
```

Perform kwash vs others analysis using conditional logistic regression at admission to determine which LV are interesting

```{r}
library(survival)
library(dplyr)

admission <- prot %>% filter(timepoint == "A")
admission$Kall <- ifelse(admission$Group == "Kwash", 1, 0)
table(admission$Kall, admission$Group)

fit <- clogit(Kall ~ LMV7 + strata(Pair), data = admission)
summary(fit)

```
Loop to do clogit in all LV's, extract estimate and FDR p-value

```{r}

GLM.run<-function(y) {
  form <- as.formula(paste0("Kall ~ strata(Pair) +", y))
  fit<-(clogit(form, data=admission))}

GLMs.out <- lapply(colnames(admission[,c(11:26)]),GLM.run )
results<-lapply(GLMs.out, function(x){summary(x)})
results

#extracting pValues and FDR correction
results<-lapply(GLMs.out, function(x){coef(summary(x))[1,5]})
pVal = as.matrix(results)
rownames(pVal) = colnames(admission[11:26])
fdr = as.matrix(p.adjust(as.vector(results), method ="fdr", n=16)) #n=number of metabolites
pVal=cbind(pVal, fdr)
colnames(pVal) = c("pVal", "fdr p")
pVal <- as.matrix(pVal)
View(as.data.frame(pVal))

#extracting estimate
results_est<-lapply(GLMs.out, function(x){coef(summary(x))[1,"exp(coef)"]}) #[x,4] change x to number of factors + 1
estimate = as.matrix(results_est)
rownames(estimate) = colnames(admission[11:26])
View(as.data.frame(estimate))

#extracting std error
results_se<-lapply(GLMs.out, function(x){coef(summary(x))[1,"se(coef)"]}) #[x,4] change x to number of factors + 1
stderror = as.matrix(results_se)
rownames(stderror) = colnames(admission[11:26])
View(as.data.frame(stderror))

#compile statistics table
uni_stat <- cbind(estimate, stderror, pVal)
colnames(uni_stat) <- c("exp(estimate)", "std error", "pVal", "FDRpVal")
View(as.data.frame(uni_stat))
write.csv(uni_stat, file = "admission_clogit_matchstrata.csv")
uni_stat
```


```{r}

# Create a boxplot for each variable in the list
library(gridExtra)
library(ggplot2)

variable_names <- colnames(admission)[-c(1:10,27)] # Exclude the 'Group' column

# Create individual boxplots
plot_list <- lapply(variable_names, function(variable) {
  ggplot(admission, aes_string(x = "Group", y = variable)) +
    geom_boxplot(aes(fill = Group), color = "black", outlier.shape = 21, outlier.size = 2, width = 0.5) +
    labs(x = "Group", y = variable) +
    theme_minimal(base_size = 15) +
    theme(legend.position = "none") +
    scale_fill_brewer(palette = "Set3")
})

# Arrange the boxplots in a 3x3 grid
grid.arrange(grobs = plot_list, ncol = 4)
```

Only LV5 and LV7 pass criterion 1

Second assessment, LV should be associated with severity of oedema

```{r}

kadmission <- admission %>% filter(Group == "Kwash")
head(kadmission)

kadmission$edema_s[kadmission$edema_s == "+"] <- 1
kadmission$edema_s[kadmission$edema_s == "++"] <- 2
kadmission$edema_s[kadmission$edema_s == "+++"] <- 3
kadmission$edema_s <- as.factor(kadmission$edema_s)

library(MASS)

ol5 = polr(edema_s ~ LMV5, data=kadmission, Hess = TRUE)
summary(ol5)
(ctable <- coef(summary(ol5)))
pval <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
(ctable <- cbind(ctable, "p value" = pval))

ol7 = polr(edema_s ~ LMV7, data=kadmission, Hess = TRUE)
summary(ol7)
(ctable <- coef(summary(ol7)))
pval <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
(ctable <- cbind(ctable, "p value" = pval))

```
Visualize it to make sure


```{r}
# Create individual boxplots
plot_list <- lapply(variable_names, function(variable) {
  ggplot(admission, aes_string(x = "edema_s", y = variable)) +
    geom_boxplot(aes(fill = edema_s), color = "black", outlier.shape = 21, outlier.size = 2, width = 0.5) +
    labs(title = variable, x = "Edema severity", y = variable) +
    theme_minimal(base_size = 15) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          legend.position = "none") +
    scale_fill_brewer(palette = "Set3")
})

# Arrange the boxplots in a 3x3 grid
grid.arrange(grobs = plot_list, ncol = 4)

```


Criterion 3, it should resolve over time

```{r}

boxplot(LMV5 ~ timepoint*Group, data =prot)
boxplot(LMV7 ~ timepoint*Group, data =prot)

library(lme4)
library(lmerTest)

resolve5 <- lmer(LMV5 ~ timepoint*Group + (1 | subjid), data = prot)
summary(resolve5)

resolve7 <- lmer(LMV7 ~ timepoint*Group + (1 | subjid), data = prot)
summary(resolve7)

table(prot$Group)

```

