---
title: "Plasma Metabolome FA analysis"
author: "Bryan Gonzales"
date: "2024-04-11"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is the analysis of the urine metabolomics data after factor analysis

```{r cars}
prot <- read.csv("FAD_urine_metabolome.csv", header = TRUE, sep = ",")
head(prot)
colnames(prot)
```
Perform PCA on entire metabolome

```{r}
pca <- prcomp(prot[,-c(1:9)])
prot$Group <- as.factor(prot$Group)
scores <- as.data.frame(pca$x)

# Combine scores with group information
pca_df <- cbind(scores, Group = prot$Group)

# Plot PCA using ggplot2
library(ggplot2)

ggplot(pca_df, aes(x = PC1, y = PC2, color = Group, label = Group)) +
  geom_point(size = 3) +
  labs(x = "PC1", y = "PC2", title = "PCA Plot") +
  theme_minimal() +
  scale_color_discrete(name = "Group", labels = c("Kwash", "Maras", "Nonmal")) # Add labels for color
```


Perform kwash vs others analysis using conditional logistic regression at admission to determine which LV are interesting

```{r}
library(survival)
library(dplyr)

prot$Kall <- ifelse(prot$Group == "Kwash", 1, 0)
table(prot$Kall, prot$Group)

fit <- clogit(Kall ~ ULMV1 + strata(Pair), data = prot)
summary(fit)

```
Loop to do clogit in all LV's, extract estimate and FDR p-value

```{r}

GLM.run<-function(y) {
  form <- as.formula(paste0("Kall ~ strata(Pair) +", y))
  fit<-(clogit(form, data=prot))}

GLMs.out <- lapply(colnames(prot[,c(10:17)]),GLM.run )
results<-lapply(GLMs.out, function(x){summary(x)})
results

#extracting pValues and FDR correction
results<-lapply(GLMs.out, function(x){coef(summary(x))[1,5]})
pVal = as.matrix(results)
rownames(pVal) = colnames(prot[,c(10:17)])
fdr = as.matrix(p.adjust(as.vector(results), method ="fdr", n=8)) #n=number of metabolites
pVal=cbind(pVal, fdr)
colnames(pVal) = c("pVal", "fdr p")
pVal <- as.matrix(pVal)
View(as.data.frame(pVal))

#extracting estimate
results_est<-lapply(GLMs.out, function(x){coef(summary(x))[1,"exp(coef)"]}) #[x,4] change x to number of factors + 1
estimate = as.matrix(results_est)
rownames(estimate) = colnames(prot[,c(10:17)])
View(as.data.frame(estimate))

#extracting std error
results_se<-lapply(GLMs.out, function(x){coef(summary(x))[1,"se(coef)"]}) #[x,4] change x to number of factors + 1
stderror = as.matrix(results_se)
rownames(stderror) = colnames(prot[,c(10:17)])
View(as.data.frame(stderror))

#compile statistics table
uni_stat <- cbind(estimate, stderror, pVal)
colnames(uni_stat) <- c("exp(estimate)", "std error", "pVal", "FDRpVal")
View(as.data.frame(uni_stat))
write.csv(uni_stat, file = "urine_clogit_matchstrata.csv")
uni_stat
```


```{r}
variable_names <- colnames(prot)[c(10:17)]

library(ggplot2)
library(gridExtra)

# Create individual boxplots
plot_list <- lapply(variable_names, function(variable) {
  ggplot(prot, aes_string(x = "Group", y = variable)) +
    geom_boxplot(aes(fill = Group), color = "black", outlier.shape = 21, outlier.size = 2, width = 0.5) +
    labs(title = variable, x = "Group", y = variable) +
    theme_minimal(base_size = 15) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          legend.position = "none") +
    scale_fill_brewer(palette = "Set3")
})

# Arrange the boxplots in a 3x3 grid
grid.arrange(grobs = plot_list, ncol = 4)

```

Only ULV1 

Second assessment, LV should be associated with severity of oedema

```{r}

prot$edema_s[prot$edema_s == "+"] <- 1
prot$edema_s[prot$edema_s == "++"] <- 2
prot$edema_s[prot$edema_s == "+++"] <- 3
prot$edema_s <- as.factor(prot$edema_s)

library(MASS)
filtered_data <- subset(prot, edema_s == 1 | edema_s == 2 | edema_s == 3)

ol1 = polr(edema_s ~ ULMV1 + age + sex + triage, data=filtered_data, Hess = TRUE)
summary(ol1)
(ctable <- coef(summary(ol1)))
pval <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
(ctable <- cbind(ctable, "p value" = pval))


```
Visualize it to make sure

```{r}
boxplot(ULMV1 ~ Group, data = prot)
boxplot(ULMV1 ~ edema_s, data = prot)

ggplot(prot, aes(x = edema_s, y = ULMV1)) +
    geom_boxplot(aes(fill = edema_s), color = "black", outlier.shape = 21, outlier.size = 2, width = 0.5) +
    labs(x = "edema_s", y = "uLV1") +
    theme_minimal(base_size = 15) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          legend.position = "none") +
    scale_fill_brewer(palette = "Set3")


```



See contribution of ULMV1 to PC1

```{r}

plot(pca_df$PC2, prot$ULMV1)

```

Regroup and check if LV1 is associated with SAM

```{r}
prot$sam <- as.factor(ifelse(prot$Group == "Nonmal", 0, 1))
table(prot$sam, prot$Group)

fit <- clogit(sam ~ ULMV1 + strata(Pair), data = prot)
summary(fit)

boxplot(ULMV1 ~ sam, prot)


ggplot(prot, aes(x = sam, y = ULMV1)) +
    geom_boxplot(aes(fill = sam), color = "black", outlier.shape = 21, outlier.size = 2, width = 0.5) +
    labs(x = "sam", y = "uLV1") +
    theme_minimal(base_size = 15) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          legend.position = "none") +
    scale_fill_brewer(palette = "Set3")

```

Run partial correlation network analysis to integrate metabolomics and clinical biochemisty

```{r}

clin <- read.csv("FAD_urine_metabolome_clinical.csv", header = TRUE, sep = ";")
head(clin)


library(psych)
library(qgraph)
library(glasso)

ADclass2 <- clin[,-c(1,8:15)]

# Calculate partial correlations
partial_cor <- cor_auto(ADclass2)

# Create network plot
qgraph(partial_cor, layout = "spring", minimum = 0.3)

sparse_partial_cor <- glasso(partial_cor, rho = 0.1)

# Create network plot
qgraph(sparse_partial_cor, layout = "spring", minimum = 0.2, labels = colnames(ADclass2))
qgraph(sparse_partial_cor, layout = "circle", minimum = 0.2, labels = colnames(ADclass2))



```


