---
title: "Factor Analysis of Kwashiorkor Lipidomics Data"
author: "Gerard Bryan Gonzales"
output: html_document
date: "2023-12-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is the analysis for the kwashiorkor MSF plasma lipidomics data using factor analysis.

Load pre-processed data

```{r}

data <- read.table("lipidome_plasma.csv", header = T, sep = ",")
head(data)[1:15]
```

We perform fad at admission

Extract latent variables from admission data only

```{r}
library(dplyr)
library(fad)

admission <- data %>% filter(timepoint == "A")

demog <- admission[,c(1:11)]
prot <- data.matrix(admission[,-c(1:11)])
head(prot)[,1:5]
```

Perform exploratory factor analysis using fad package.

First is to optimize factors to obtain a factor with least BIC

```{r}
library(fad)

BICs <- rep(0,101) # first optimize factors ranging from 1 to 100

for(i in 1:101){
  out = fad(x = prot,factors = i-1, rotation = "promax") # x is the highly dimensional dataset where p>n
  BICs[i] = out$BIC
}

BICs
plot(BICs)

ind = which.min(BICs) # obtain the optimal factor favored with lowest BIC 

ind
```

Based on the BIC, the optimal number of factors is 23. Hence, we run factor analysis using 9 factors

```{r}

out1 = fad(x = prot, factors = 23, scores = c("Bartlett"), rotation = "varimax", lower = 0.01) ## X is the data
eig <- data.frame(out1$scores)
head(eig)

kdat <- cbind(demog, eig)
head(kdat)

write.csv(kdat, "factor_scores.csv", row.names = F)
```

Extract the loadings from FAD. Loadings are the contributions of each protein on the latent variable.
```{r}

loadings <- data.matrix(out1$loadings)
head(loadings)

lipids <- colnames(prot)

rownames(loadings) <- lipids

```

We impose a factor loading cut-off. we will use 0.40. This means that proteins that have less than 0.4 loadings to the latent variable will be removed - leading behind only those proteins that are contributing highly to the latent variable.

We then perform confirmatory factor analysis using lavaan using only strongly associated proteins per latent variable.

Then, we extract a new scores matrix - the predicted factor so that each observation gets a value for each latent variable.

```{r}

load_cut <- loadings
load_cut[load_cut < abs(0.4)] <- 0
head(load_cut)

write.csv(load_cut, "loadings_lipidomeFAD.csv")
```

Extract scores using the load_cut. We removed LVs without any loading (LV 21 and 23)

```{r}
library(psych)

load_cut <- load_cut[,-c(21,23)]

fxad <- factor.scores(prot, load_cut, Phi = NULL, method = c("Bartlett"),rho=NULL,impute="none")

scores_ad <- fxad$scores
head(scores_ad)

CFA_lipidome_admission <- cbind(demog, scores_ad)

colnames(CFA_lipidome_admission)[12:32] <- paste0("lLV", 1:21)
head(CFA_lipidome_admission)

write.csv(CFA_lipidome_admission, file= "CFA_lipidome_admission.csv", row.names=F)

```


Do the discharge and treated timepoints

```{r}
###Do discharge

discharge <- data %>% filter(timepoint == "D")
demog_d <- discharge[,c(1:11)]
prot_d <- discharge[,-c(1:11)]

fxd <- factor.scores(prot_d, load_cut, Phi = NULL, method = c("Bartlett"),rho=NULL,impute="none")

scores_d <- fxd$scores
head(scores_d)

CFA_lipidome_discharge <- cbind(demog_d, scores_d)
colnames(CFA_lipidome_discharge)[12:32] <- paste0("lLV", 1:21)
head(CFA_lipidome_discharge)

write.csv(CFA_lipidome_discharge, file= "CFA_lipidome_discharge.csv", row.names=F)

###do treated

treated <- data %>% filter(timepoint == "TRUE")
demog_t <- treated[,c(1:11)]
prot_t <- treated[,-c(1:11)]

fxt <- factor.scores(prot_t, load_cut, Phi = NULL, method = c("Bartlett"),rho=NULL,impute="none")

scores_t <- fxt$scores
head(scores_t)

CFA_lipidome_treated <- cbind(demog_t, scores_t)
colnames(CFA_lipidome_treated)[12:32] <- paste0("lLV", 1:21)
head(CFA_lipidome_treated)

write.csv(CFA_lipidome_treated, file= "CFA_lipidome_treated.csv", row.names=F)

###combine

CFA_lipidome <- rbind(CFA_lipidome_admission, CFA_lipidome_discharge, CFA_lipidome_treated)
write.csv(CFA_lipidome, file= "CFA_lipidome_all.csv", row.names=F)

```

Annotate the proteins

```{r}
#just to check the loadings

annot <- read.csv("Lipidomics_Plasma_annot.csv", header = T, sep =";")
loadings <- read.csv("loadings_lipidomeFAD.csv", header = T, sep =",")
head(loadings)

colnames(loadings)[1] <- "Code"
head(loadings)

loadings <- merge(loadings, annot, by = "Code")
head(loadings)

write.csv(loadings, file = "loadings_lipidomeFAD.csv", row.names = F)

```

Try out some latent variables. Visualise difference among groups, across edema grades and resolution over time

```{r}
boxplot(CFA_lipidome_admission$lLV7 ~ CFA_lipidome_admission$Group)
boxplot(CFA_lipidome_admission$lLV7 ~ CFA_lipidome_admission$edema_s)
boxplot(CFA_lipidome$lLV7 ~ CFA_lipidome$Group*CFA_lipidome$timepoint)
boxplot(CFA_lipidome$lLV7 ~ CFA_lipidome$timepoint*CFA_lipidome$Group)

boxplot(CFA_lipidome_admission$lLV3 ~ CFA_lipidome_admission$Group)
boxplot(CFA_lipidome_admission$lLV3 ~ CFA_lipidome_admission$edema_s)
boxplot(CFA_lipidome$lLV3 ~ CFA_lipidome$Group*CFA_lipidome$timepoint)
boxplot(CFA_lipidome$lLV3 ~ CFA_lipidome$timepoint*CFA_lipidome$Group)

```


Correlate individual proteins with LV

```{r}


non_zero_codes <- loadings$Code[loadings$V5 != 0]
non_zero_codes

CFAlipids <- admission[,c(non_zero_codes)]

#replace name

name_df <- data.frame(
  old_name = loadings$Code,
  new_name = loadings$Lipid
)
name_mapping <- setNames(name_df$new_name, name_df$old_name)

colnames(CFAlipids) <- name_mapping[colnames(CFAlipids)]

head(CFAlipids)

library(ggplot2)
library(reshape2)
library(Hmisc)

heat <- cbind(CFA_lipidome_admission$lLV5, CFAlipids)
colnames(heat)[1] <- "lLV5"

# Calculate the correlation matrix and p-values using the Hmisc package
cor_results <- rcorr(as.matrix(heat))
cor_matrix <- cor_results$r
p_matrix <- cor_results$P

# Melt the correlation and p-value matrices
cor_melted <- melt(cor_matrix)
p_melted <- melt(p_matrix)
colnames(cor_melted) <- c("Var1", "Var2", "Correlation")
colnames(p_melted) <- c("Var1", "Var2", "p_value")

# Merge the melted dataframes
cor_p_data <- merge(cor_melted, p_melted, by = c("Var1", "Var2"))

# Create the heatmap with both correlation coefficients and p-values
w <- ggplot(data = cor_p_data, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f\np=%s", Correlation, ifelse(p_value < 1e-10, "<1E-10", sprintf("%.3g", p_value)))), color = "black", size = 4) +
  scale_fill_gradient2(low = "red", high = "blue", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       name="Correlation") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                   size = 12, hjust = 1)) +
  coord_fixed() +
  labs(title = "Heatmap of Correlation Matrix with P-values", x = "", y = "")

custom_colors <- colorRampPalette(c("white", "blue"))(100)

w <- ggplot(data = cor_p_data, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile() +  scale_fill_gradientn(colors = custom_colors) +
    theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                   size = 12, hjust = 1)) +
  coord_fixed()
  
```


