---
title: "Factor Analysis Metabololics Kwash"
author: "Bryan Gonzales"
date: "2024-01-02"
output: html_document
---

This is the analysis for the kwashiorkor MSF proteomics data using factor analysis.

Load pre-processed data
```{r}
data <- read.table("MSCombine_plasma_metabolomics.csv", header = T, sep = ",")
head(data)[1:15]
colnames(data)[1:15]
```
We perform CFA per time point.

Extract latent variables from admission data only

```{r}
library(dplyr)

admission <- data %>% filter(timepoint == "A")

demog <- admission[,c(1:10)]
prot <- data.matrix(admission[,-c(1:10)])
head(prot)[,1:5]
```
Perform exploratory factor analysis using fad package.

First is to optimize factors to obtain a factor with least BIC

```{r}
library(fad)

BICs <- rep(1,101)

suppressWarnings({ for(i in 1:101){
  out = fad(x = prot,factors = i-1, rotation = "varimax", lower = 0.005)
  BICs[i] = out$BIC
}
})

BICs
plot(BICs)

ind = which.min(BICs) # obtain the optimal factor favored with lowest BIC 

ind
```
Based on the BIC, the optimal number of factors is 16. Hence, we run factor analysis using 16 factors

```{r}

out1 = fad(x = prot, factors = 16, scores = c("Bartlett"), rotation = "varimax", lower = 0.01)
eig <- data.frame(out1$scores)
head(eig)

kdat <- cbind(demog, eig)
head(kdat)

write.csv(kdat, "factor_scores.csv", row.names = F)
```

Extract the loadings from FAD. Loadings are the contributions of each protein on the latent variable.
```{r}

loadings <- data.matrix(out1$loadings)
head(loadings)

proteins <- colnames(prot)

rownames(loadings) <- proteins

```

We impose a factor loading cut-off. we will use 0.3. This means that proteins that have less than 0.3 loadings to the latent variable will be removed - leading behind only those proteins that are contributing highly to the latent variable.

We then perform confirmatory factor analysis using lavaan using only strongly associated proteins per latent variable.

Then, we extract a new scores matrix - the predicted factor so that each observation gets a value for each latent variable.

```{r}

load_cut <- loadings
load_cut[load_cut < abs(0.3)] <- 0
head(load_cut)

write.csv(load_cut, "loadings_metabolomeFAD.csv")
```

Extract scores using the load_cut

```{r}
library(psych)

fxad <- factor.scores(prot, load_cut, Phi = NULL, method = c("Bartlett"),rho=NULL,impute="none")

scores_ad <- fxad$scores
head(scores_ad)

CFA_proteome_admission <- cbind(demog, scores_ad)

colnames(CFA_proteome_admission)[11:26] <- c("LMV1", "LMV2", "LMV3", "LMV4", "LMV5", "LMV6", "LMV7", "LMV8", "LMV9",
                                             "LMV10", "LMV11", "LMV12", "LMV13", "LMV14", "LMV15", "LMV16")
head(CFA_proteome_admission)

write.csv(CFA_proteome_admission, file= "CFA_metabolome_admission.csv", row.names=F)

```

Do the discharge and treated timepoints

```{r}
###Do discharge

discharge <- data %>% filter(timepoint == "D")
demog_d <- discharge[,c(1:10)]
prot_d <- discharge[,-c(1:10)]

fxd <- factor.scores(prot_d, load_cut, Phi = NULL, method = c("Bartlett"),rho=NULL,impute="none")

scores_d <- fxd$scores
head(scores_d)

CFA_proteome_discharge <- cbind(demog_d, scores_d)
colnames(CFA_proteome_discharge)[11:26] <- c("LMV1", "LMV2", "LMV3", "LMV4", "LMV5", "LMV6", "LMV7", "LMV8", "LMV9",
                                             "LMV10", "LMV11", "LMV12", "LMV13", "LMV14", "LMV15", "LMV16")
head(CFA_proteome_discharge)

write.csv(CFA_proteome_discharge, file= "CFA_proteome_discharge.csv", row.names=F)

###do treated

treated <- data %>% filter(timepoint == "T")
demog_t <- treated[,c(1:10)]
prot_t <- treated[,-c(1:10)]

fxt <- factor.scores(prot_t, load_cut, Phi = NULL, method = c("Bartlett"),rho=NULL,impute="none")

scores_t <- fxt$scores
head(scores_t)

CFA_proteome_treated <- cbind(demog_t, scores_t)
colnames(CFA_proteome_treated)[11:26] <- c("LMV1", "LMV2", "LMV3", "LMV4", "LMV5", "LMV6", "LMV7", "LMV8", "LMV9",
                                             "LMV10", "LMV11", "LMV12", "LMV13", "LMV14", "LMV15", "LMV16")
head(CFA_proteome_treated)

write.csv(CFA_proteome_treated, file= "CFA_proteome_treated.csv", row.names=F)

###combine

CFA_proteome <- rbind(CFA_proteome_admission, CFA_proteome_discharge, CFA_proteome_treated)
write.csv(CFA_proteome, file= "CFA_proteome_all.csv", row.names=F)

```

Try out some latent variables. Visualise difference among groups, across edema grades and resolution over time

```{r}
boxplot(CFA_proteome_admission$LMV7 ~ CFA_proteome_admission$Group)
boxplot(CFA_proteome_admission$LMV7 ~ CFA_proteome_admission$edema_s)
boxplot(CFA_proteome$LMV7 ~ CFA_proteome$Group*CFA_proteome$timepoint)
boxplot(CFA_proteome$LMV7 ~ CFA_proteome$timepoint*CFA_proteome$Group)


```
